Multi-state modelling with msm: a practical course

### 2021-09-23 docs/msmcourse_src/02-msm-basic.Rmd
Minor fixes for course ch1-2

  - Time should be increasing within each patient

@@

Instead, draw a diagram of allowed _continuous-time_ transitions, and choose `Q` to describe the structure of this diagram.

@@

If the intensities $q_{rs}$ are constant over time, then $E(T_r) = -1/q_{rr}$ where $q_{rr} = -\sum_{s!=r} q_{rs}$

@@

* 10 months in state 1 before moving to state 2 : $q_{12} = 0.1$

* 10 months in state 2 before moving to state 3 : $q_{23} = 0.1$

* 10 months in state 3 before moving to state 4 : $q_{23} = 0.1$

@@

Here is an artificial example: 
```{r,echo=FALSE}
cav_obstype <- cav[1:7,c("PTNUM","years","state")]
cav_obstype$obstype <- c(1,1,1,2,1,2,3)
cav_obstype
```

@@

3. From the May workshop notes, remind yourself of how the transition intensities are related to the probabilities governing the state that someone will move to on leaving the current state.

    a. From your data summary, make a guess at what these might be.

@@

    Note that there is no "right answer" for these initial guesses.  The learning objective is to be able to convert between transition intensities and more easily-understood quantities.

@@

1. 
    ```{r}
    statetable.msm(state, PTNUM, data=cav)
    ```

@@

    We then set the remaining, off-diagonal values $q_{rs}, s!=r$ according to guesses of the probability $\pi_{rs}$ that the next state will be $s$, and deducing that since $\pi_{rs} = $-q_{rs} / q_{rr}$, we can initialise $q_{rs} = \pi_{rs} T_r$.

>>>>

  - Time should be increasing within each patient (but first observation doesn't have to be at time 0)

@@

Instead, it should be based on scientific judgement about what happens in continuous time.

@@

If the intensities $q_{rs}$ are constant over time, then

* $E(T_r) = -1/q_{rr}$ where $q_{rr} = -\sum_{s!=r} q_{rs}$, i.e.

* the sum of intensities in row $r$ (excluding the diagonal $q_{rr}$) is 1 / the mean sojourn time in state $r$

@@

* 10 months in state 1 before moving to state 2 : $-1/q_{11} = 10$, hence $q_{12} = 0.1$, since in this model we can only transition to $s=2$ from $r=1$, and similarly,

* 10 months in state 2 before moving to state 3 : $q_{23} = -q_{22} = 0.1$

* 10 months in state 3 before moving to state 4 : $q_{34} = -q_{33} = 0.1$

@@

Here is an artificial dataset, where we know the patient:

* was in state 1 at times 0 and 1 (`obstype` 1)

* occupied state 2 throughout the period between years 2 and 3.09 and between years 4 and 5 (`obstype` 2)

* died at time 5.85 (`obstype` 3)

```{r,echo=FALSE}
cav_obstype <- cav[1:7,c("PTNUM","years","state")]
cav_obstype$obstype <- c(1,1,1,2,1,2,3)
cav_obstype
```

@@

3. In the CAV model, there are two alternative "next states" for people in state 1 and people in state 2.   From the [May workshop notes](https://chjackson.github.io/msm/msmcourse/cjackson_bordeaux2020.pdf), remind yourself of how the transition intensities are related to the probabilities governing the state that someone will move to on leaving the current state.

    a. From your data summary, make a guess at what these probablities might be.

@@

    Note there is no single "right answer" for these initial guesses.  The learning objective is to be able to convert between transition intensities and more easily-understood quantities.

@@

1. 
    ```{r}
    library(msm)
    statetable.msm(state, PTNUM, data=cav)
    ```

@@

    We then set the remaining, off-diagonal values $q_{rs}, s!=r$ according to guesses of the probability $\pi_{rs}$ that the next state will be $s$, and deducing that since $\pi_{rs} = -q_{rs} / q_{rr}$, we can initialise $q_{rs} = \pi_{rs} / T_r$.
